Information Distribution Depends on Language-Specific Features

### 2019-01-30 cogsci/position-info.Rmd
cogsci full draft minus Yurovsky edits

Our replication of the @yu2016 results

@@

@@

Our plots for three representative utterance lengths are below.

@@

@@

After replicating the @yu2016 results using the spoken conversation corpus from @levy2007, we want to determine whether the entropy curves will remain robust over child-adult conversational corpora.

The corpora in the CHILDES database consists of short and often disconnected utterances across hours of recordings, the unigram entropy measure is unaffected by context or lack thereof in utterances by the adults and children in the corpora. For our first corpus, we used the Providence English corpus from CHILDES @macwhinney2014. The Providence corpus recorded interactions between children between 1 and 3 years old and their parents in the home. We divided the corpus by speaker into child and non-child categories. We further divided the corpus by utterance length, so that all sentences of length $k$ (e.g. $6$) were grouped together. Finally, within each utterance length, we computed the unigram entropy measure for each position. Computation of positional entropy was identical to our general method after dividing for utterance length.

We also ran this analysis on Spanish and Mandarin corpora from CHILDES. We used the Shiro corpus for Spanish @shiro2000, which contains prompted narratives individually collected from over a hundred Venezualan schoolchildren, half from high SES backgrounds and half from low SES backgrounds. We used the Zhou dinner corpus for Mandarin Chinese (Li & Zhou, 2015), which contains dinner conversations between 5 to 6-year-old children and their parents collected in Shanghai.

@@

@@

We found a distinct three-step distribution for English and Spanish CHILDES corpora, with a slight dip in the penultimate position of each sentence. The final position of utterances in child-directed speech is known to be important, dating back to Aslin (1993). The Mandarin corpus entropy curve, by comparison, has a noticeably lower positional entropy values in utterance-final positions than in utterance-penultimate positions.

We attribute the penultimate dip in the English and Spanish entropy curves to the fact that most of the utterances in the CHILDES English and Spanish corpora we examined had a determiner such as "the" or "a" in the second-to-last position of utterances. The beginnings of utterances in the English and Spanish CHILDES data were usually pronouns or grammatical subjects, while the final words were grammatical objects and had a great deal of variation in the exact word that appeared in the utterance-final position.

These entropy curves are not what would be expected from UID. The robustness of the three-step distribution for English and its replication in Spanish do not resemble the affine function we would expect from UID. The Mandarin entropy curve, which does not at all resemble either the English/Spanish distribution or the predictions of UID, suggests that the entropy curve can vary from language to language. UID predicts that each language should have a similar distribution.

@@

However, sentences of length greater than ten in the Wikipedia corpora were significantly more common than in the CHILDES. Sentences on Wikipedia are commonly , then we also computed relative slope treatments. In the *relative 5* treatment, we computed the slopes between $0\%$ and $20\%$, $20\%$ and $40\%$, $40\%$ and $60\%$, $60\%$ and $80\%$ and $80\%$ and $100\%$ of the relative word positions in each sentence. When one of these percentages was not a whole number, then the closest whole number position was used instead for slope calculation. Each comparative slope within each treatment was averaged together between different sentence lengths, for example in the relative $5$ treatment then all of the $0\%$ to $20\%$ slopes were averaged together. This created three treatments for the entropy curve for each langauge in the Wikipedia database.

In both the absolute and relative $5$ treatments, each language is embedded in $5$-dimensional space. This allows for direct comparison between languages on Wikipedia using cosine similarity and moreover for unsupervised clustering analysis to compare the results of our position-wise analysis on datasets from Wikipedia to known typological features within the languages. To determine which phonological, morphological and syntactic features affected the embedding of a language in the Wikipedia dataset, we use the linguistic features in World Atlas of Language Structures (Dryer & Haspelmath, 2013).

We check the effects of individual features on the embeddings of languages in the different treatments. We computed pairwise cosine similarity between each pair of language vectors within a treatment. For a subset of WALS features which had values entered for most of the languages we obtained from Wikipedia, we used a generalized linear model to see whether the cosine similarity between languages mattered in predicting if the languages shared the same value for a WALS feature.

@@

For eight features and $45$ languages, we found this to be true. The table below shows the results for the generalized linear model we computed using WALS features and the absolute treatment.

@@

Cosine distance played some role in the feature determination. The top four features all characterize word order, which indicates that for this subset of features and languages that the embeddings in the slope space are related to the WALS features. Therefore typological features play a role in determining the positional entropy values for a language. This means that the entropy curves are in some way structured by the syntactic, morphological and phonological features of a language.

@@

An ideal approach would use all of the corpora we can download from Wikipedia and all $144$ features from the WALS database. Cluster the languages used in the Wikipedia analysis on the basis of WALS features and then directly compare the WALS clustering results with the results of the hierarchical clustering on the Wikipedia slope data for each of the different treatments using clustering similarity evaluation methods such as the Rand index (Rand, 1971). However, the problem then arises of which combination of WALS features and how many features to include in the clustering analysis. This is a computationally intractable operation. Additionallly, deciding how many clusters to use in an unsupervised clustering analysis is an unsolved problem in machine learning.

On a more mechanical level, we expect that our entropy model will have implications for how long people take to read individual words at different positions of a sentence. Eye-tracking research has indicated the effects of surprisal on fixation duration during eye-tracking studies [@demberg2008; @boston2008; @smith2013]; higher surprisal of a word is on average predictive of higher fixation duration. The *wrap-up effect* states that when a person reads written text, he or she will process sentence-final words more slowly on average then sentence-medial or sentence-initial words, due to integrating information from the entire sentence to form a final understanding of the sentence's meaning [@stowe2018; @kuperman2010]. The wrap-up effect is drawn from evidence in languages with a large final increase in their entropy curve from our study, so perhaps the wrap-up effect derives from the sentence-final increase in entropy curves observed in English and other Germanic languages.

>>>>

We replicated the @yu2016 results and found the same characteristic three-step distribution as they did for the BNC entropy curves.

@@

The positional entropy at the beginning of sentences is low, then rises and plateaus for sentence-medial word positions before dropping slightly in the second-to-last position and rising again. However, this curve could simply be a result of the written words in the BNC. We will expand this analysis and replicate the result with a spoken American English corpus.

@@

Our plots for three representative utterance lengths are below. We found the same characteristic shape for the American English Switchboard corpus as in our replication of the @yu2016 results for the BNC.

@@

The shape we find in these two English data is notable for a number of reasons, but what we find most interesting is one, the shape of the distribution does not follow our predictions from UID and two, the distribution is robust across an unrelated pair of spoken and written English corpora. This suggests that the entropy curve is characteristic of the English language at least, and is not a specific effect of computing positional entropies in the written British English texts in the BNC. For our next step, we want to expand the positional entropy analysis to child-adult conversational corpora and languages beyond English to see if the same three-step distribution appears.

@@

After replicating the @yu2016 results using the spoken conversation corpus from @levy2007, we want to determine whether the entropy curves will remain robust over child-adult conversational corpora. We also wanted to see whether the curves would show the same characteristic three-step shape across languages. To accomplish both of these goals, we used data from the CHILDES Talkbank system @macwhinney2014.

The corpora in the CHILDES database consists of short and often disconnected utterances across hours of recordings, the unigram entropy measure is unaffected by context or lack thereof in utterances by the adults and children in the corpora. For our first corpus, we used the Providence English corpus @demuth2006word. The Providence corpus recorded interactions between children between 1 and 3 years old and their parents in the home. We divided the corpus by speaker into child and non-child categories. We further divided the corpus by utterance length, so that all sentences of length $k$ (e.g. $6$) were grouped together. Finally, within each utterance length, we computed the unigram entropy measure for each position. Computation of positional entropy was identical to our general method after dividing for utterance length.

We also ran this analysis on Spanish and Mandarin corpora from CHILDES. Spanish is an Indo-European language like English, possessing similar grammar, word order and numerous cognate word. Mandarin Chinese is typologically completed unrelated to English. We used the Shiro corpus for Spanish @shiro2000, which contains prompted narratives individually collected from over a hundred Venezualan schoolchildren, half from high SES backgrounds and half from low SES backgrounds. We used the Zhou dinner corpus for Mandarin Chinese @zhou2015dinner, which contains dinner conversations between 5 to 6-year-old children and their parents collected in Shanghai.

@@

We expect that our entropy analysis on the Providence corpus will produce an identical entropy curve shape to the BNC and Switchboard curves. We expect similar results to English in the Spanish Shiro entropy curve, while we expect that our Mandarin corpus will produce a different entropy curve shape.

@@

We found a distinct three-step distribution for English and Spanish CHILDES corpora, with a slight dip in the penultimate position of each sentence. The Mandarin corpus entropy curve, by comparison, has a noticeably lower positional entropy values in utterance-final positions than in utterance-penultimate positions. We attribute the penultimate dip in the English and Spanish entropy curves to the fact that most of the utterances in the CHILDES English and Spanish corpora we examined had a determiner such as "the" or "a" in the second-to-last position of utterances. The beginnings of utterances in the English and Spanish CHILDES data were usually pronouns or grammatical subjects, while the final words were grammatical objects and had a great deal of variation in the exact word that appeared in the utterance-final position.

These entropy curves are not what would be expected from UID. The robustness of the three-step distribution for English and its replication in Spanish do not resemble the affine function we would expect from UID. The Mandarin entropy curve, which does not at all resemble either the English/Spanish distribution or the predictions of UID, suggests that the entropy curve can vary from language to language. UID predicts that each language should have a similar distribution, which should approximate an affine function.

We found that all of the English corpora followed the same three-step entropy curve shape, regardless of whether the corpus contained spoken or written data. The Spanish corpus we used also followed this same shape, which we expected. However, Mandarin did not have the same shape, indicating that the shape of the entropy curves may arise from cross-linguistic variation. Similar grammatical and other features between two languages, such as Spanish and English, may produce a similar entropy curve.

@@

However, sentences of length greater than ten in the Wikipedia corpora were significantly more common than in the CHILDES. Therefore we also computed relative slope treatments. In the *relative 5* treatment, we computed the slopes between every $20%$ of the relative word positions in each sentence length, e.g. the slope between $0\%$ and $20\%$ of sentences of length $10$ is the slope between the position entropies at the sentence-initial word position and the position representing the third word. When computing the word position index to make those cuts, if the slope was not a whole number, then the closest whole number position was used instead for slope calculation. Each comparative slope within each treatment was averaged together between different sentence lengths, for example all of the $0\%$ to $20\%$ slopes over all sentence lengths were averaged together.

In both the absolute and relative $5$ treatments, each language is embedded in $5$-dimensional space. This permitted us to use cosine similarity for direct comparison between the languages we pulled from Wikipedia. To determine which phonological, morphological and syntactic features affected the embedding of a language in the Wikipedia dataset, we use the linguistic features in World Atlas of Language Structures @dryer2013wals. We checked the effects of individual features on the embeddings of languages in the different treatments. We computed pairwise cosine similarity between each pair of language vectors within each treatment. For a subset of eight WALS features, we used a generalized linear model to see whether the cosine similarity between languages mattered in predicting if the languages shared the same value for a WALS feature.

@@

For eight features and $45$ languages we pulled from Wikipedia, we found that the cosine similarity between WALS features. The table below shows the results for the generalized linear model we computed using WALS features and the absolute treatment.

@@

From these results, we argue that cosine distance played some role in the feature determination. The top four features in determining the WALS feature value all characterize word order, which indicates that for this subset of features and languages that the embeddings in the slope space are related to the WALS features. We argue therefore that typological features play a role in determining the positional entropy values for a language. This indicates that the entropy curves are in some way structured by the syntactic, morphological and phonological features of a language.

@@

We initially wanted to use several hundred language corpora pulled from Wikipedia and all $144$ linguistic features from WALS. We considered an computing unsupervized clusterings of the $5$-dimensional vectors of our Wikipedia entropy curves, and comparing those clusterings to clusterings of the WALS features directly using a cluster similarity measure such as the Rand index @rand1971objective. However, the problem then arises of which combination of WALS features and how many features to include in the clustering analysis, with $144!$ different combinations. Two additional problems presented themselves: one, deciding how many clusters to use in an unsupervised clustering analysis is an unsolved problem in machine learning; and two, not all of the languages on Wikipedia have values inputted for their WALS features. As a follow-up, we are considering missing-data imputation on the WALS features in order to run the generalized linear model analysis on a large dataset of languages.

On a more mechanical level, we expect that our entropy model will have implications for how long people take to read individual words at different positions of a sentence. Eye-tracking research has indicated the effects of surprisal on fixation duration during eye-tracking studies [@demberg2008; @boston2008; @smith2013]; higher surprisal of a word is on average predictive of higher fixation duration. The *wrap-up effect* states that people process sentence-final words more slowly on average then sentence-medial or sentence-initial words when reading, due to integrating information to form a final understanding of the sentence's meaning [@stowe2018; @kuperman2010]. The wrap-up effect is drawn from evidence in languages with a large final increase in their entropy curve from our study, so we hypothesize that the supposed wrap-up effect derives from the same source sentence-final increase in entropy curves observed in English and other Germanic languages.
